---
title: "First summary of chemostat model"
author: "Daniel Schwartz"
date: "September 6, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#Clear environment
rm(list=ls())
pd=par()
#Load necessary library
library(deSolve)
library(tidyverse)
```

We have discussed the idea of testing whether phages can prevent the loss of sporulation. The idea discussed was using a chemostat since it has been reported that spo- mutants can overtake chemostat populations within a couple of weeks (Dawes and Mandelstam 1970). Could the addition of phage result in maintenance of the sporulation trait during continuous culture?
Before setting up the experiment I decide to run simulations of such chemostat experiments. A basic chemostat model of bacteria (**N**) growing in a limiting resource (**R**) is described by the following set of ODEs:

(@) $\frac{dR}{dt}=D(Rin-R)-\mu RN$

(@) $\frac{dN}{dt}=\mu RN - DN$

```{r , echo=FALSE, message=FALSE, fig.align='center'}
chemostat <- function(t,current,params){
  with(as.list(c(params,current)), {
    #list the equations for the derivatives of each state variable
    #R is the resource
    dR <- D * (R_in - R) - mu * R * N 
    #N are the bacteria 
    dN <- mu * R * N - D * N 
    
    results <- c(dR, dN)
    return (list(results))
  })
}

###define parameters
# resource in inflow
R_in <- 1e6
#dilution rate
D <- 0.1
# growth rate
mu <- D
# all parameters together
params <- c(R_in=R_in, D=D, mu=mu)

#declare initial conditions for state variables
current0 <- c(R = R_in, N = 100)

#define the time to run model
t <- seq (0, 100, by=1)

#solve the model and save the output
out <- lsoda(current0, t, chemostat, params)

# transform to long format for ggplot
out.l <- gather(as.data.frame(out), key='population', value='Conc', colnames(out)[-1])
# To savve time I will save the results of the first run and read results from file next times
write_csv(out.l, "./summary1_output/out_01.csv")
out.l <- read_csv("./summary1_output/out_01.csv")
#plot
ggplot(out.l, aes(x=time, y=Conc))+
  geom_line(aes(color=population),size=2)+
  scale_y_log10()+
  annotation_logticks(sides = 'l')+
  theme_bw()+
  scale_colour_brewer(palette='Set1')+
  ggtitle(paste(
    paste0(paste(names(params)),'=',paste(params), collapse = '; '),'\n',
    paste0(paste(names(current0)),'=',paste(current0), collapse = '; ')))

```

# Adding Spores to the System
First lets add sporulation as a single step process. In this model a constant fraction of cells ($\alpha$)  sporulate into spores (**S**) and a constant fraction of the spores ($\beta$) resucitate back into cells.

(@) $\frac{dR}{dt}=D(Rin-R)-\mu RN$

(@) $\frac{dN}{dt}=\mu RN -\alpha N+\beta S-  DN$

(@) $\frac{dS}{dt}=\alpha N-\beta S - DS$

```{r , echo=FALSE, message=FALSE, fig.align='center'}
chemostat1 <- function(t,current,params){
  with(as.list(c(params,current)), {
    #list the equations for the derivatives of each state variable
    #R is the resource
    dR <- D * (R_in - R) - mu * R * N 
    
    #N are the bacteria 
    dN <- mu * R * N - D * N - alpha * N + beta * S
    #S are the spores 
    dS <- alpha * N - beta * S - D*S
    
    results <- c(dR, dN, dS)
    return (list(results))
  })
}

###define parameters
# resource in inflow
R_in <- 1e6
#dilution rate
D <- 0.1
# growth rate
mu <- D
# probaility of sporulation
alpha <- 0.01
#probability of reactivation of spores
beta <- 0.01

# all parameters together
params <- c(R_in=R_in, D=D, mu=mu, alpha=alpha, beta=beta)

#declare initial conditions for state variables
current1 <- c(R = R_in, N = 100, S=1)


#define the time to run model
t <- seq (0, 100, by=1)

#solve the model and save the output
out <- lsoda(current1, t, chemostat1, params)

# transform to long format for ggplot
out.l <- gather(as.data.frame(out), key='population', value='Conc', colnames(out)[-1])
# To save time I will save the results of the first run and read results from file next times
write_csv(out.l, "./summary1_output/out_02.csv")
out.l <- read_csv("./summary1_output/out_02.csv")
#plot
ggplot(out.l, aes(x=time, y=Conc))+
  geom_line(aes(color=population), size=2)+
  scale_y_log10()+
  annotation_logticks(sides = 'l')+
  theme_bw()+
  scale_colour_brewer(palette='Set1')+
  ggtitle(paste(
    paste0(paste(names(params)),'=',paste(params), collapse = '; '),'\n',
    paste0(paste(names(current1)),'=',paste(current1), collapse = '; ')))

```

We can see that sporulation reduces the steady state value of the vegetative cells bellow the limiting resource value. As a consequence there is also a sight increase in steady state resource value.
In Bacillus subtilis sporulation in fact takes much longer than vegetative cell replication: sporulation is on the order of hours and replication about 1/2 hour. To reflect these time dicrepencies I model sporulatiion a a chain of transitions from cell to spore through several intermediate stages. Each of these transitions takes one time unit and thus the maturation to a spore lasts as long as the number of intermediate steps. In the following example we sporulation takes 3 stages (2 intermediates (**I1,I2**)and mature spore)

(@) $\frac{dR}{dt}=D(Rin-R)-\mu RN$

(@) $\frac{dN}{dt}=\mu RN -\alpha N+\beta S-  DN$

(@) $\frac{dI1}{dt}=\alpha N -I1 - DI1$

(@) $\frac{I2}{dt}=I1 -I2 - DI2$

(@) $\frac{dS}{dt}=I2-\beta S - DS$

Note that each of the sporulation stages had a dilution loss term. The result is that only some fraction of cells that initate sporulation will contribute to the spore population.

```{r , echo=FALSE, message=FALSE, fig.align='center'}
chemostat3 <- function(t,current,params){
  with(as.list(c(params,current)), {
    #list the equations for the derivatives of each state variable
    #R is the resource
    dR <- D * (R_in - R) - mu * R * N 
    
    #N are the bacteria 
    dN <- mu * R * N - D * N - alpha * N + beta * S
        # Sporulation takes 3 time steps. modeled as 3 population maturing through the process
    dI1 <- alpha * N - I1 - D*I1
    dI2 <- I1  - I2 - D*I2
    #S are the spores 
    dS <- I2 - beta * S - D*S
    
    results <- c(dR, dN, dI1, dI2, dS)
    return (list(results))
  })
}


#declare initial conditions for state variables
current3 <- c(R = R_in, N = 100, I1=1, I2=1,S=1)
#define the time to run model
t <- seq (0, 100, by=1)

#solve the model and save the output
out <- lsoda(current3, t, chemostat3, params)

# transform to long format for ggplot
out.l <- gather(as.data.frame(out), key='population', value='Conc', colnames(out)[-1])
# To save time I will save the results of the first run and read results from file next times
write_csv(out.l, "./summary1_output/out_03.csv")
out.l <- read_csv("./summary1_output/out_03.csv")
out.l$population <- as.factor(out.l$population)
out.l$population <-factor(out.l$population, levels = c('N', 'R', 'S','I1','I2'))

#plot
ggplot(out.l, aes(x=time, y=Conc))+
  geom_line(aes(color=population), size=2)+
  scale_y_log10()+
  annotation_logticks(sides = 'l')+
  theme_bw()+
  scale_colour_brewer(palette='Set1')+
  ggtitle(paste(
    paste0(paste(names(params)),'=',paste(params), collapse = '; '),'\n',
    paste0(paste(names(current3)),'=',paste(current3), collapse = '; ')))

```

We expect that the more stages there are to sporulation the less spores there will be at stedy state.

```{r , echo=FALSE, message=FALSE, fig.align='center'}
chemostat6 <- function(t,current,params){
  with(as.list(c(params,current)), {
    #list the equations for the derivatives of each state variable
    #R is the resource
    dR <- D * (R_in - R) - mu * R * N 
    
    #N are the bacteria 
    dN <- mu * R * N - D * N - alpha * N + beta * S
    
    # Sporulation takes 6 time steps. modeled as 6 population maturig through the process
    dI1 <- alpha * N - I1 - D*I1
    dI2 <- I1  - I2 - D*I2
    dI3 <- I2 - I3 - D*I3
    dI4 <- I3 - I4 - D*I4
    dI5 <- I4 - I5 - D*I5
    dS <- I5 - beta * S - D*S
    
    results <- c(dR, dN, dI1, dI2,dI3,dI4,dI5, dS)
    return (list(results))
  })
}
chemostat9 <- function(t,current,params){
  with(as.list(c(params,current)), {
    #list the equations for the derivatives of each state variable
    #R is the resource
    dR <- D * (R_in - R) - mu * R * N 
    
    #N are the bacteria 
    dN <- mu * R * N - D * N - alpha * N + beta * S
    
    # Sporulation takes 9 time steps. modeled as 9 population maturing through the process
    dI1 <- alpha * N - I1 - D*I1
    dI2 <- I1 - I2 - D*I2
    dI3 <- I2 - I3 - D*I3
    dI4 <- I3 - I4 - D*I4
    dI5 <- I4 - I5 - D*I5
    dI6 <- I5 - I6 - D*I6
    dI7 <- I6 - I7 - D*I7
    dI8 <- I7 - I8 - D*I8
    dS  <- I8 - beta * S - D*S
    
    results <- c(dR, dN, dI1, dI2,dI3,dI4,dI5,dI6, dI7, dI8, dS)
    return (list(results))
  })
}

current6 <- c(R = R_in, N = 100, I1=1, I2=1,I3=1, I4=1,I5=1,S=1)
current9 <- c(R = R_in, N = 100, I1=1, I2=1,I3=1, I4=1,I5=1,I6=1, I7=1,I8=1,S=1)

out <- lsoda(current6, t, chemostat6, params)
# transform to long format for ggplot
out.l <- gather(as.data.frame(out), key='population', value='Conc', colnames(out)[-1])
# To save time I will save the results of the first run and read results from file next times
write_csv(out.l, "./summary1_output/out_04.csv")
out.l <- read_csv("./summary1_output/out_04.csv")
out.l$population <- as.factor(out.l$population)
out.l$population <-factor(out.l$population, levels = c('N', 'R', 'S','I1','I2','I3','I4','I5'))

#plot
ggplot(out.l, aes(x=time, y=Conc))+
  geom_line(aes(color=population), size=2)+
  scale_y_log10()+
  annotation_logticks(sides = 'l')+
  theme_bw()+
  scale_colour_brewer(palette='Set1')+
  ggtitle(paste(
    paste0(paste(names(params)),'=',paste(params), collapse = '; '),'\n',
    paste0(paste(names(current6)),'=',paste(current6), collapse = '; ')))


out <- lsoda(current9, t, chemostat9, params)
# transform to long format for ggplot
out.l <- gather(as.data.frame(out), key='population', value='Conc', colnames(out)[-1])
# To save time I will save the results of the first run and read results from file next times
write_csv(out.l, "./summary1_output/out_05.csv")
out.l <- read_csv("./summary1_output/out_05.csv")
out.l$population <- as.factor(out.l$population)
out.l$population <-factor(out.l$population, levels = c('N', 'R', 'S','I1','I2','I3','I4','I5','I6','I7','I8'))

#plot
ggplot(out.l, aes(x=time, y=Conc))+
  geom_line(aes(color=population), size=2)+
  scale_y_log10()+
  annotation_logticks(sides = 'l')+
  theme_bw()+
  scale_colour_brewer(palette='Set1')+
  ggtitle(paste(
    paste0(paste(names(params)),'=',paste(params), collapse = '; '),'\n',
    paste0(paste(names(current9)),'=',paste(current9), collapse = '; ')))



```

Now we can put the results of the models with different stages of sporulation side by side to see the diference:

```{r , echo=FALSE, message=FALSE, fig.align='center'}
out.l <- read_csv("./summary1_output/out_01.csv")
out.l$stages <- 0
out.all <- out.l
out.l <- read_csv("./summary1_output/out_02.csv")
out.l$stages <- 1
out.all <- rbind(out.all,out.l)
out.l <- read_csv("./summary1_output/out_03.csv")
out.l$stages <- 3
out.all <- rbind(out.all,out.l)
out.l <- read_csv("./summary1_output/out_04.csv")
out.l$stages <- 6
out.all <- rbind(out.all,out.l)
out.l <- read_csv("./summary1_output/out_05.csv")
out.l$stages <- 9
out.all <- rbind(out.all,out.l)

out.all <- filter(out.all, population=='N'|population=='R'|population=='S')

out.l$population <- as.factor(out.l$population)
out.l$population <-factor(out.l$population, levels = c('N', 'R', 'S'))

ggplot(out.all, aes(x=time, y=Conc))+
  geom_line(aes(color=population), size=2)+
  scale_y_log10()+
  theme_bw()+
  scale_colour_brewer(palette='Set1')+
  facet_grid(.~stages, scales = "free_x")
```

The number of stges does not seem to have a large influence. Lets overlay the populations from the different mosdels:

```{r , echo=FALSE, message=FALSE, fig.align='center'}

ggplot(out.all, aes(x=time, y=Conc))+
  geom_line(aes(color=as.character(stages)), size=2)+
  scale_y_log10()+
  theme_bw()+
  scale_colour_brewer(palette='Set1')+
  facet_grid(.~population, scales = "free_x")

```

As expected, the longer it takes to sporulate the less spores you get at steady state.

##Dilution rate and time length of sporulation
How do dilution rates and the time it takes to complete sporulation combine to determine the density of spores in the chemostat model?

```{r , echo=FALSE, message=FALSE, fig.align='center'}
t <- seq (0, 500, by=1)


d.rates <- data.frame(dilution=rep(c(0.01,0.03,0.06,0.09,0.15,0.3,0.6,0.9,1),4),
                      spo.time= as.numeric(rep(c(1,3,6,9),each=9)),ratio=NA, N=NA, S=NA)

for (i in 1:nrow(d.rates)){
  #dilution rate
  D <- d.rates$dilution[i]
  mu <- D
  # all parameters together
  params <- c(R_in, D, mu, alpha, beta)
  t.spo <- d.rates$spo.time[i]
  if (t.spo==1){
    current1 <- c(R = R_in, N = 100, S=1)
    out <- lsoda(current1, t, chemostat1, params)

  } else if (t.spo==3){
    current3 <- c(R = R_in, N = 100, I1=1, I2=1,S=1)
    out <- lsoda(current3, t, chemostat3, params)

  } else if (t.spo==6){
    current6 <- c(R = R_in, N = 100, I1=1, I2=1,I3=1, I4=1,I5=1,S=1)
    out <- lsoda(current6, t, chemostat6, params)

  } else if (t.spo==9){
    current9 <- c(R = R_in, N = 100, I1=1, I2=1,I3=1, I4=1,I5=1,I6=1, I7=1,I8=1,S=1)
    out <- lsoda(current9, t, chemostat9, params)

  }

  rel <- as.data.frame(out)
  rel$all <- apply(X = rel[,3:ncol(rel)],MARGIN = 1,FUN = sum)
  rel$rel <- rel$S/rel$all
  d.rates$ratio[i] <- rel$rel[nrow(rel)]
  d.rates$N[i] <- rel$N[nrow(rel)]
  d.rates$S[i] <- rel$S[nrow(rel)]
}
write_csv(d.rates, "./summary1_output/out_06.csv")
d.rates <- read_csv("./summary1_output/out_06.csv")
p1 <- 
  ggplot(d.rates, aes(x=dilution, y=ratio))+
  geom_line(aes(color=as.factor(spo.time)),size=1)+
  geom_point(aes(shape=as.factor(spo.time),color=as.factor(spo.time)),size=3)+
  scale_color_discrete(name="Spor.\ntime")+
  scale_shape_discrete(name="Spor.\ntime")+
  scale_y_log10(breaks=c(1e-1,1e-2,1e-3,1e-4))+
  ylab("spore frequency (log)")+
  xlab("Dilution rate (hr^-1)")+ theme_bw()+
  theme(legend.position = "bottom")+
  annotation_logticks(sides = 'l')+
  guides(col = guide_legend(nrow = 2))+
  ggtitle("spore frequency")

p2 <- 
  ggplot(d.rates, aes(x=dilution, y=S))+
  geom_line(aes(color=as.factor(spo.time)),size=1)+
  geom_point(aes(shape=as.factor(spo.time),color=as.factor(spo.time)),size=3)+
  scale_color_discrete(name="Spor.\ntime")+
  scale_shape_discrete(name="Spor.\ntime")+
  scale_y_log10(limits=c(1e1,1e6))+
  ylab("Spore conc. (log)")+
  xlab("Dilution rate (hr^-1)")+ theme_bw()+
  theme(legend.position = "bottom")+
  annotation_logticks(sides = 'l')+
  guides(col = guide_legend(nrow = 2))+
  ggtitle("Spore concentration")
p3 <- 
  ggplot(d.rates, aes(x=dilution, y=N))+
  geom_line(aes(color=as.factor(spo.time)),size=1)+
  geom_point(aes(shape=as.factor(spo.time),color=as.factor(spo.time)),size=3)+
  scale_color_discrete(name="Spor.\ntime")+
  scale_shape_discrete(name="Spor.\ntime")+
  # scale_y_log10(limits=c(1e1,1e6))+ 
  ylab("Veg. cell conc. ")+
  xlab("Dilution rate (hr^-1)")+ theme_bw()+
  theme(legend.position = "bottom")+
  # annotation_logticks(sides = 'l')+
  guides(col = guide_legend(nrow = 2))+
  ggtitle("Veg. cell concentration")

library(gridExtra)
grid.arrange(p1,p2,p3,nrow=1)


```

As the dilution rate increases the time it takes to complete sporulation becomes a more important factor in determining the steady state level of spores. Note however, that this has no effect on the actual concnetration of the cells. Cell densities do vary somewhat with dilution rate (this plot is on linear y).

##Loss of sporulaton
In the next model a mutation enables the loss of sporulation. 

(@) $\frac{dR}{dt}=D(Rin-R)-\mu R(N+N_{mut})$

(@) $\frac{dN}{dt}=\mu RN -\alpha N+\beta S-\theta N -   DN$

(@) $\frac{dS}{dt}=\alpha N-\beta S - DS$

(@) $\frac{dN_{mut}}{dt}=\mu RN_{mut}+\theta N - DN_{mut}$

Beside sporulation there is no change in parameters of the mutant compared with the WT sporulator. The model is the single step sporulation model (no intermediates). the mutation rate to lose sporulation ($\delta$) is 10^-6^. 

```{r , echo=FALSE, message=FALSE, fig.align='center'}
chemostat1.mut <- function(t,current,params){
  with(as.list(c(params,current)), {
    #list the equations for the derivatives of each state variable
    #R is the resource
    dR <- D * (R_in - R) - mu * R * (N+Nmut) 
    
    #N are the sporulating bacteria 
    dN <- mu * R * N - D * N - alpha * N + beta * S - theta * N
    # Nmut are the spo- mutants
    dNmut <- mu * R * Nmut - D*Nmut +theta * N
    
    #S are the spores 
    dS <- alpha * N - beta * S - D*S
    
    
    results <- c(dR, dN,dNmut, dS)
    return (list(results))
  })
}
#dilution rate
D <- 0.1
# growth rate
mu <- D

#loss of sporulation by mutation
theta <- 1e-6

# all parameters together
params1 <- c(R_in=R_in, D=D, mu=mu, alpha=alpha, beta=beta,theta=theta)
#declare initial conditions for state variables
current1 <- c(R = R_in, N = 1e2, Nmut=0, S=0)
#define the time to run model
t <- seq (0, 3000, by=1)

#solve the model and save the output
out <- lsoda(current1, t, chemostat1.mut, params1)
out.l <- gather(as.data.frame(out), key='population', value='Conc', colnames(out)[-1])
# # To save time I will save the results of the first run and read results from file next times
write.csv(out.l, "./summary1_output/out_07.csv")
out.l <- read_csv("./summary1_output/out_07.csv")
out.l <- out.l[,-1]
out.l$population <- as.factor(out.l$population)
out.l$population <-factor(out.l$population, levels = c('N', 'R', 'S','Nmut'))


# ggplot(out.l[out.l$population!='R',], aes(x=time, y=Conc))+
ggplot(out.l, aes(x=time, y=Conc))+
  geom_line(aes(color=population), size=2)+
  scale_y_log10()+
  theme_bw()+
  scale_colour_brewer(palette='Set1')+
  ggtitle(paste(
    paste0(paste(names(params1)),'=',paste(params1), collapse = '; '),'\n',
    paste0(paste(names(current1)),'=',paste(current1), collapse = '; ')))
```

The sporulator is outcompeted by the mutant!
(warning message is beacuse the system was intiated with Nmut=0 and S=0, which doesn't work well on log scale...)


##Adding a virus into the chemostat

First thing I will ad a virus into the basic chemostst model. without sporulation.


(@) $\frac{dR}{dt}=D(Rin-R)-\mu RN$

(@) $\frac{dN}{dt}=\mu RN -\delta NV - DN$

(@) $\frac{dV}{dt}=\delta NV (\phi-1)- DN$

```{r , echo=FALSE, message=FALSE, fig.align='center'}
chemostat0v <- function(t,current,params){
  with(as.list(c(params,current)), {
    #list the equations for the derivatives of each state variable
    #R is the resource
    dR <- D * (R_in - R) - mu * R * N 
    
    #N are the bacteria 
    dN <- mu*R * N - delta * N * V  -  D*N

    #V are the phages
    dV <- delta * N * V * (phi-1) - D*V
    results <- c(dR, dN, dV)
    return (list(results))
  })
}

#adsorption constant rate (per hour per ml)
delta <- 1e-8

# busrt size
phi <- 50

params0v <- c(R_in=R_in, D=D, mu=mu,  delta=delta, phi= phi)
#declare initial conditions for state variables
current0v <- c(R = R_in, N = 100,  V=1)
#define the time to run model
t <- seq (0, 100, by=1)
#run the model
out <- lsoda(current0v, t, chemostat0v, params0v)

out.l <- gather(as.data.frame(out), key='population', value='Conc', colnames(out)[-1])
# # To save time I will save the results of the first run and read results from file next times
write_csv(out.l, "./summary1_output/out_08.csv")
out.l <- read_csv("./summary1_output/out_08.csv")
#add empty value for S to work mae colors consistent
out.l <- rbind(out.l, data.frame(time=0, population='S', Conc=NA))
out.l$population <- as.factor(out.l$population)
out.l$population <-factor(out.l$population, levels = c('N', 'R', 'S','V'))


# ggplot(out.l[out.l$population!='R',], aes(x=time, y=Conc))+
ggplot(out.l, aes(x=time, y=Conc))+
  geom_line(aes(color=population), size=2)+
  scale_y_log10()+
  theme_bw()+
  scale_colour_brewer(palette='Set1')+
  ggtitle(paste(
    paste0(paste(names(params0v)),'=',paste(params0v), collapse = '; '),'\n',
    paste0(paste(names(current0v)),'=',paste(current0v), collapse = '; ')))

```

The addition of the virus results in a viral limited population rather than a resource-limited population.
Next we'll combine the virus model and the sporulation model. Here only active cells are suceptible to viral infection. We will use the single stage sporulation model.


(@) $\frac{dR}{dt}=D(Rin-R)-\mu RN$

(@) $\frac{dN}{dt}=\mu RN -\alpha N+\beta S-\delta NV-  DN$

(@) $\frac{dS}{dt}=\alpha N-\beta S - DS$

(@) $\frac{dV}{dt}=\delta NV (\phi-1)- DN$


```{r , echo=FALSE, message=FALSE, fig.align='center'}
chemostat1v <- function(t,current,params){
  with(as.list(c(params,current)), {
    #list the equations for the derivatives of each state variable
    #R is the resource
    dR <- D * (R_in - R) - mu * R * N 
    
    #N are the bacteria 
    dN <- mu*R * N + alpha * S - D * N - beta * N - delta * N * V
    #S are the spores 
    dS <- alpha * N - beta * S - D*S
    
    #V are the phages
    dV <- delta * N * V * (phi-1) - D*V
    results <- c(dR, dN, dS, dV)
    return (list(results))
  })
}

params1v <- c(R_in=R_in, D=D, mu=mu, alpha=alpha, beta=beta, delta=delta, phi=phi)
current1v <- c(R = R_in, N = 100, S=1, V=1)
#define the time to run model
t <- seq (0, 100, by=1)
#run the model
out <- lsoda(current1v, t, chemostat1v, params1v)

out.l <- gather(as.data.frame(out), key='population', value='Conc', colnames(out)[-1])
# # To save time I will save the results of the first run and read results from file next times
write_csv(out.l, "./summary1_output/out_09.csv")
out.l <- read_csv("./summary1_output/out_09.csv")
out.l$population <- as.factor(out.l$population)
out.l$population <-factor(out.l$population, levels = c('N', 'R', 'S','V'))


# ggplot(out.l[out.l$population!='R',], aes(x=time, y=Conc))+
ggplot(out.l, aes(x=time, y=Conc))+
  geom_line(aes(color=population), size=2)+
  scale_y_log10()+
  theme_bw()+
  scale_colour_brewer(palette='Set1')+
  ggtitle(paste(
    paste0(paste(names(params1v)),'=',paste(params1v), collapse = '; '),'\n',
    paste0(paste(names(current1v)),'=',paste(current1v), collapse = '; ')))



```

Now we can compare the reults of the sporulation-chemostat with and without sporulation and viruses


```{r , echo=FALSE, message=FALSE, fig.align='center'}
out.noV <- read_csv("./summary1_output/out_02.csv")
out.noV$model <- "Spores"
out.noS <- read_csv("./summary1_output/out_08.csv")
out.noS$model <- "Virus"
out.basic <- read_csv("./summary1_output/out_01.csv")
out.basic$model <- "Basic"
out.l$model <- "Spores+Virus"
out.l <- rbind( out.l[-nrow(out.l),],out.noV,out.noS, out.basic)
out.l$population <- as.factor(out.l$population)
out.l$population <-factor(out.l$population, levels = c('N', 'R', 'S','V'))

ggplot(out.l, aes(x=time, y=Conc))+
  geom_line(aes(color=population), size=2)+
  scale_y_log10()+
  theme_bw()+
  scale_colour_brewer(palette='Set1')+
  facet_grid(.~model, scales = "free_x")

```

The ability to sporulate results in a population that is not at the max density set by the resource, similar to the addition of a virus (though less dramatic). My interpertation of this is that spotulation, like infection, is a costly lost term.
There are also some subtle differences in the time it takes for the system to reach steady state. This will be easier to see if we put all on the same time scale:

```{r , echo=FALSE, message=FALSE, fig.align='center'}

ggplot(out.l, aes(x=time, y=Conc))+
  geom_line(aes(color=population), size=2)+
  scale_y_log10()+
  theme_bw()+
  scale_colour_brewer(palette='Set1')+
  facet_grid(model~., scales = "fixed")

```