---
title: "Serial tranfer model with sporulation"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::knit_hooks$set(inline = function(x) {
  knitr:::format_sci(x, 'md')
})
#Clear environment
rm(list=ls())
pd=par()
#Load necessary library
library(deSolve)
library(tidyverse)

setwd("~/GitHub/chemostat/code/serial_ transfer/")
```
The chemostat model seems unhelpful to test the idea that viruses can prevent the loss of sporulation. As an alternative I will try modeling a serial transfer experiment to test if this experimental setup is more suitable.

I will be implementing the serial transfer model of:

Chaudhry, W. N., Pleška, M., Shah, N. N., Weiss, H., McCall, I. C., Meyer, J. R., … Levin, B. R. 
(2018). Leaky resistance and the conditions for the existence of lytic bacteriophage. 
PLOS Biology, 16(8), e2005971. https://doi.org/10.1371/journal.pbio.2005971 

# Simple model
I will start with a model of serial transfer of a bacterium without sorulation and without phage. The model tracks changes in resouces (**R**) and bacteria (**N**). 
According to Chandhury et al. growth dependence on resouces is given as a Monod function:

$$ \psi(R) =  \frac {R}{k+R}$$
 
 "*where R is the concentration (μg per mL) of the limiting resource and k is the “Monod constant,” corresponding to the concentration of the resource (μg per mL), at which the bacteria grow at a half of their maximum rate.*"

Stripping there equations from phages and mutants, and adding a mortality term (**m**) for cells I am left with

(@) $$\frac{dR}{dt}=-e\cdot \psi(R)\cdot vN $$

(@) $$ \frac{dN}{dt}=\psi(R)\cdot v \cdot N - m\cdot N$$

where **e** is *a conversion efficacy parameter (μg per cell)* and **v** is the *maximum growth rate*.
```{r}
#Define system of differential equations for resource and bacteria
serial <- function(t,current,params){
  with(as.list(c(params,current)), {
    #list the equations for the derivatives of each state variable
    #R is the resource (μg per mL)
    dR =  -e *(R/(R+k))*v*N  
    
    #N are the bacteria 
    dN = (R/(R+k))*v*N -m*N


    results <- c(dR, dN )
    return (list(results))
  })
}

###define parameters
# maximum growth rate
v <- 0.2
#conversion efficacy parameter
e <- 5e-7
#Half-saturation constant of resource uptake by bacteria
k = 1
#mortality
m=0.01
# all parameters together
params <- c (v=v,e=e, k=k,m=m)

#declare initial conditions for state variables
R_t0 <- 10
N_t0 <- 1e5
current <- c(R = R_t0 , N = N_t0)

#define the time to run model
t <- seq (0, 24, by=1)

# run the model
out <- lsoda(current, t, serial, params)
# plot(out)
#transform to long format for ggplot
out.l <- gather(as.data.frame(out), key='population', value='Conc', colnames(out)[-1])

#plot
ggplot(out.l, aes(x=time, y=Conc))+
  geom_line(size=1)+
  theme_bw()+
  scale_colour_brewer(palette='Set1')+
  ggtitle(paste(
    paste0(paste(names(params)),'=',paste(params), collapse = '; '),'\n',
    paste0(paste(names(current)),'=',paste(current), collapse = '; ')))+
  facet_grid(population~., scales = "free_y")

```

#### adding serial transfers
I will now use the events functionality of the desolve package to replenish nutrients every 24 hours
```{r}

# add nutrients and dilute culture at regular time intervals
event <- function(t, current, parms) {
  current['R'] <- R_t0
  current['N'] <- 0.1 * current['N']
  return(current)
}

#define the time to run model
t <- seq (0, 240, by=1)
#deine time of transfer
t.transfer <- 24
t.event <- t[t%%t.transfer==0][-1] #[-1] removes t=0

out <- lsoda(current, t, serial, params,
             events = list(func = event, time=t.event))

# transform to long format for ggplot
out.l <- gather(as.data.frame(out), key='population', value='Conc', colnames(out)[-1])

#plot
ggplot(out.l, aes(x=time, y=Conc))+
  geom_line(size=1)+
  theme_bw()+
  scale_colour_brewer(palette='Set1')+
  ggtitle(paste(
    paste0(paste(names(params)),'=',paste(params), collapse = '; '),'\n',
    paste0(paste(names(current)),'=',paste(current), collapse = '; ')))+
  facet_grid(population~., scales = "free_y")
  
```

#### Adding sporulation
I need to add terms that will make sporulation (**s**) and germiation (**g**) sependent on **R**. I came up with the following functions:


$$ s(R)=2^{-R} $$
$$g(R)= \frac {R}{R+1} $$

```{r}
R <- 0:10#seq(0,1000,by = 1)
par(mfrow=c(2, 1) )
plot(R, 2^-(0.5*R), main="sporulation", type='l', col="red")
plot(R, R/(R+.1), main="germination", type='l', col="blue")

```

These functions are quite abrupt but they will have to do for now. I will start with a simple model without transfers.
```{r}
#Define system of differential equations for resource and bacteria
serial <- function(t,current,params){
  with(as.list(c(params,current)), {
    #list the equations for the derivatives of each state variable
    #R is the resource (μg per mL)
    dR =  -e *(R/(R+k))*v*N  
    
    #N are the bacteria 
    dN = (R/(R+k))*v*N +(R/(R+.1))*S -(2^(-R/2))*N -m*N
   
    #S are the spores 
    dS = (2^(-R/2))*N -(R/(R+.1))*S

    results <- c(dR, dN, dS )
    return (list(results))
  })
}

###define parameters
N_t0 <- 100
S_t0 <- 1
current <- c(R = R_t0 , N = N_t0, S=S_t0)

#define the time to run model
t <- seq (0, 240, by=1)

# run the model
out <- lsoda(current, t, serial, params)
# plot(out)
# transform to long format for ggplot
out.l <- gather(as.data.frame(out), key='population', value='Conc', colnames(out)[-1])

#plot
ggplot(out.l, aes(x=time, y=Conc))+
  geom_line(size=1)+
  theme_bw()+
  scale_colour_brewer(palette='Set1')+
  ggtitle(paste(
    paste0(paste(names(params)),'=',paste(params), collapse = '; '),'\n',
    paste0(paste(names(current)),'=',paste(current), collapse = '; ')))+
  facet_grid(population~., scales = "free_y")
```

That seems to be working. I will next att serial transfers to the sporulation model

```{r}

# add nutrients and dilute culture at regular time intervals
event <- function(t, current, parms) {
  current['R'] <- R_t0
  current['N'] <- 0.1 * current['N']
  return(current)
}

#define the time to run model
t <- seq (0, 1480, by=1)
#deine time of transfer
t.transfer <-100
t.event <- t[t%%t.transfer==0][-1] #[-1] removes t=0

out <- lsoda(current, t, serial, params,
             events = list(func = event, time=t.event))

# transform to long format for ggplot
out.l <- gather(as.data.frame(out), key='population', value='Conc', colnames(out)[-1])

#plot
ggplot(out.l, aes(x=time, y=Conc))+
  geom_line(size=1)+
  theme_bw()+
  # scale_y_log10()+
  scale_colour_brewer(palette='Set1')+
  ggtitle(paste(
    paste0(paste(names(params)),'=',paste(params), collapse = '; '),'\n',
    paste0(paste(names(current)),'=',paste(current), collapse = '; ')))+
  facet_grid(population~., scales = "free_y")
  
```

I need to figure out why the population size is growing indefinetly. I may have run int the atto-fox pronlem again :(